### 2025-09-18 paper/sec/introduction.tex

\section{Introduction: Transparent, Data-Driven Documents}

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain, because of the difficulty in tracing specific claims back to the data. Peer review of empirical claims
without author-supplied traceability to the underlying data can be challenging~\citep{weber20}; paper
retractions are often attributable to (non-fraudulent but significant) errors in data management and
analysis~\citep{hu25}. The use of LLMs to assist with these challenges has received considerable attention
recently, on problems ranging from fact-checking~\citep{abu-ahmad25} to interpretation of charts and
figures~\citep{roberts24}, LLM interfaces do not support direct interrogation of visual outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} approaches the
transparency problem from this direction, through software infrastructure which transparently connections
charts and figures to their data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.}
This enables visual outputs to provide interactive \emph{provenance queries}, user interactions (e.g. mousing
over a visual element) that allow readers to explore how visual features relate to data. The value proposition
of these approaches is that the interactions are provided automatically via trusted infrastructure, namely an
underlying programming language (say query language or general-purpose language) which tracks dependencies as
data flows through a computation. However, these approaches are limited to outputs computed from data, which
usually precludes natural language, despite the fact that the main claims of a paper are usually made using
natural language, with charts and figures only playing a supporting role.

In this paper, we make the case for combining these two approaches, leveraging the ability of LLMs to
understand complex technical natural language and to synthesise programs that express the analyses that
underwrite those claims, and extend interactive provenance queries to the natural language that accompanies
the charts and figures.  ..language which also makes quantitative statements and contains substantial implicit
analytic content. In principle transparent programming languages can also be used to create natural language
which is bidirectly linked to data, by computing key fragments of natural language from data. The following
example, adapted from \cite{zhang18} via \cite{moosavi21}, shows how the language presented in this paper can
be used to achieve this \rpnote{Need to illustrate counterfactual/generalisability}:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example, showing data-driven natural language (top
    vs.~bottom)}\label{fig:scigen-example-website}
\end{figure}

@@

\subsection{Target idioms of natural language}

NLP aspect of the problem is potentially a big problem space in itself. We will restrict interest to certain
idiomatic uses of natural language in making/justifying scientific claims. Table~\ref{tab:fluid_examples}

>>>>

\section{Introduction: Transparent Documents}

Making sense of and verifying data-driven claims is hard, even when the evidence base (data sets and related
analyses) is open. We see this in peer review, misinformation, retracted papers, \rpnote{supporting
citations}. The ability of LLMs to assist with these challenges has received considerable attention recently,
for example in quantity extraction~\citep{bolucu23}, scientific fact-checking~\citep{abu-ahmad25}, and
scientific figure interpretation~\citep{roberts24}.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} attempts to
address this issue through software infrastructure which transparently connections charts and figures to their
data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.} This enables visual outputs
to provide interactive \emph{provenance queries}, user interactions (e.g. mousing over a visual element) that
allow readers to explore how visual features relate to data. The key value proposition of these approaches is
that the interactions are provided automatically via an underlying programming language (query language or
general-purpose language) which tracks dependencies as data flows through a computation.

In this paper, we extend interactive provenance queries to the natural language that accompanies the charts
and figures. We believe this is an important step because in most articles, charts and figures usually only
play a supporting role. The main claims are usually presented in natural language, language which also makes
quantitative statements and contains substantial implicit analytic content. In principle transparent
programming languages can also be used to create natural language which is bidirectly linked to data, by
computing key fragments of natural language from data. The following example, adapted from \cite{zhang18} via
\cite{moosavi21}, shows how the language presented in this paper can be used to achieve this \rpnote{Need to
illustrate counterfactual/generalisability}:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example (top); counterfactual behaviour (bottom)}\label{fig:scigen-example-website}
\end{figure}

@@

\subsubsection{Target idioms of natural language}

NLP aspect of the problem is potentially a big problem space in itself. We will restrict interest to certain
idiomatic uses of natural language in making/justifying scientific claims.
Examples \ref{tab:fluid_examples}
